Multi-target detection with application to cryo-electron microscopy

### 2019-03-04 inverse_problems/inverse_problems.tex
middle of section 2 (in the model bias paragraph)

\section{Motivation: single-particle reconstruction using cryo-electron microscopy}
Cryo-electron microscopy (cryo-EM)  has recently joined X-ray crystallography and nuclear magnetic resonance (NMR) spectroscopy as a high-resolution structural method for biological macromolecules~\cite{frank2006three,kuhlbrandt2014resolution,bartesaghi20152}. 
In contrast X-ray and NMR which aggregate information from an ensembles of
particles, single particle cryo-EM produces images of individual particles and thus can, in principle, elucidate multiple  structures simultaneously.
%In addition, it does not require the formation of crystalline arrays of macromolecules.

In a cryo--EM experiment, biological samples (e.g., macromolecules, viruses) are rapidly frozen in a thin layer of vitreous ice.
The microscope produces 2-D tomographic images of the samples embedded in the ice, called \emph{micrographs}. Each micrograph contains multiple tomographic projections of the samples at unknown locations and under unknown viewing directions. 
Importantly, the electron dose must be kept low to mitigate  radiation damage, leading to high noise levels.
The goal is to construct 3-D models of the molecular structures from the micrographs.

A 2-D micrograph $y$ can be written in terms 
of the models described in Section~\ref{sec:intro} as
\begin{equation}  \label{eq:model_cryo}
y  =   s \ast x + \varepsilon,
\end{equation}
where tomographic projections are placed according to the signal $s$. In this model, the signal $x$ represents the 2-D tomographic projections of the 3-D structure taken from  random viewing directions according to (possibly unknown) distribution. In particular, by letting $V$ describes the underlying volume, $R_\omega$ a random 3-D rotation drawn from some unknown distribution and $P$ a tomographic projection, we have
\begin{equation}
x= P(R_\omega V).
\end{equation}
By allowing $V$ to be a random signal as well, this model can encode a mixture of structures  
(which might be continuous as well).

All contemporary methods in the field split the reconstruction procedure into two main  stages.
The first stage, called \emph{particle picking},  extracts the particle projections from the micrographs. Given the projections, the second stage aims to construct a 3-D model of the molecular structure, usually using a  expectation-maximization algorithm~\cite{scheres2012relion}. 
%The quality of the reconstruction eventually hinges on the quality of the particle picking stage.
Crucially, reliable detection of individual particles is impossible in a highly noisy environment. This fact has been recognized early on by the cryo-EM community. 
Particularly, in~\cite{henderson1995limitations,glaeser1999electron}, it was established that particle picking is impossible for molecules below a certain weight (below $\sim$50 kDa). 
%The unique issues raised by small particles have been mitigated by recent technological advances in the field, including the use of Volta phase plates~\cite{khoshouei2017cryo,liang2017phase} and scaffolding cages~\cite{liu2018nearatomic}.
%Despite this progress, detecting small molecules in the micrographs remains a challenge.

Another potential pitfall of particle picking is related  to \emph{model bias}, whose importance in cryo-EM was stressed by a number of authors~\cite{shatsky2009method,vanheel1992correlation,henderson2013avoiding,vanheel2013finding}. In the classical ``Einstein from noise'' experiment, multiple realizations of pure noise are aligned to a picture of Einstein using template matching and then averaged. It was shown that the averaged noise rapidly becomes remarkably similar to the Einstein template~\cite{shatsky2009method}. In the context of cryo--EM, this experiment exemplifies how prior assumptions about the particles biases the reconstructed structure. %This model bias is common to all particle picking methods based on template matching. %In our approach, no templates or human intervention are required, thus significantly reducing concerns about model bias. %

>>>>

\section{Motivation: single-particle reconstruction using cryo-electron microscopy}

Cryo-electron microscopy (cryo-EM)  has recently joined X-ray crystallography and nuclear magnetic resonance (NMR) spectroscopy as a high-resolution structural method for biological macromolecules~\cite{frank2006three,kuhlbrandt2014resolution,bartesaghi20152}.
%In contrast to X-ray and NMR which aggregate information from an ensembles of
%particles, single particle cryo-EM produces images of individual particles and thus can, in principle, elucidate multiple  structures simultaneously.
%In addition, it does not require the formation of crystalline arrays of macromolecules.
In a cryo--EM experiment, biological samples (e.g., macromolecules, viruses) are rapidly frozen in a thin layer of vitreous ice.
The microscope produces 2-D tomographic images of the samples embedded in the ice, called \emph{micrographs}. Each micrograph contains multiple tomographic projections of the samples at unknown locations and under unknown viewing directions. 
Importantly, the electron dose must be kept low to mitigate  radiation damage, inducing high noise levels.
The goal is to construct 3-D models of the molecular structures from the micrographs. 
Since  cryo-EM produces images of individual particles, it can, in principle, elucidate multiple  structures simultaneously.
This is in clear contrast to  X-ray and NMR, which aggregate information from an ensembles of
particles.

A 2-D micrograph $y$ can be written in terms 
of the models described in Section~\ref{sec:intro} as
\begin{equation}  \label{eq:model_cryo}
y  =   s \ast x + \varepsilon,
\end{equation}
where tomographic projections are placed according to the signal $s$. In this model, the signal $x$ represents 2-D tomographic projections of the 3-D structure taken from  random viewing directions according to (possibly unknown) distribution over $SO(3)$ (the group of 3-D rotations). In particular, by letting $V$ describes the underlying volume, $R_\omega$ a random 3-D rotation for $\omega\in SO(3)$, and $P$ a tomographic projection, we have
\begin{equation}
x= P(R_\omega V).
\end{equation}
By allowing $V$ to be a random signal as well, this model can encode a mixture of structures  
(which might be continuous as well).

All contemporary methods for single particle reconstruction using cryo-EM  split the reconstruction procedure into two main  stages.
The first stage, called \emph{particle picking},  detects and extracts the particle projections from the micrographs. Given the projections, the second stage aims to construct a 3-D model of the molecular structure, usually using an expectation-maximization algorithm~\cite{scheres2012relion}. 
%The quality of the reconstruction eventually hinges on the quality of the particle picking stage.
Crucially, reliable detection of individual particles is impossible in a highly noisy environment. This fact has been recognized early on by the cryo-EM community. 
Particularly, in~\cite{henderson1995limitations,glaeser1999electron}, it was established that particle picking is impossible for molecules below a certain weight (below~$\sim$50 kDa). 
%The unique issues raised by small particles have been mitigated by recent technological advances in the field, including the use of Volta phase plates~\cite{khoshouei2017cryo,liang2017phase} and scaffolding cages~\cite{liu2018nearatomic}.
%Despite this progress, detecting small molecules in the micrographs remains a challenge.

Even if particle picking is feasible, it may suffer from a phenomenon called \emph{model bias}.
The main mechanism leads most particle pickers is based on correlating the micrograph with several templates; areas with high correlation are identified as areas that  are likely to contain a projection (these are the picked particles). 
Clearly, the picked particles depend heavily on the chosen templates.
The potential pitfall of model biased
has been recognized early by the cryo-EM community~\cite{shatsky2009method,vanheel1992correlation,henderson2013avoiding,vanheel2013finding}. In the classical ``Einstein from noise'' experiment, multiple realizations of pure noise are aligned to a picture of Einstein using template matching and then averaged. It was shown that the averaged noise rapidly becomes remarkably similar to the Einstein template~\cite{shatsky2009method}. In the context of cryo--EM, this experiment exemplifies how prior assumptions about the particles biases the reconstructed structure. %This model bias is common to all particle picking methods based on template matching. %In our approach, no templates or human intervention are required, thus significantly reducing concerns about model bias. %
